---
title: "Practical Machine Learning Course Project"
output: html_document
---

# Synopsis

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. 
These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. 
They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 
More information is available from the website [here](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).

### Data

The training data for this project are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv).

The test data are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

The data for this project come from [this source](http://groupware.les.inf.puc-rio.br/har). 

### Project Goals

The goal of the project is to predict the manner in which they did the exercise. 
This is the "classe" variable in the training set. 
We may use any of the other variables to predict with.
We should create a report describing how we built the model, how we used cross validation, what we think the expected out of sample error is, and why we made the choices we did.
We will also use the prediction model to predict 20 different test cases.

# Libraries Loading

First of all we load all the libraries that we need.
```{r, message=FALSE}
library(caret)
library(ggplot2)
```

We also seet the random seed for reproducibility
```{r}
set.seed(12345)
```

# Getting the data

We download the data (if needed) and load them into memory.
```{r, cache=TRUE}
if (!file.exists("pml-training.csv")) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                destfile = "pml-training.csv")
}
if (!file.exists("pml-testing.csv")) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                destfile = "pml-testing.csv")
}

training <- read.csv("pml-training.csv", sep = ",", na.strings = c("", "NA"))
testing <- read.csv("pml-testing.csv", sep = ",", na.strings = c("", "NA"))

```

# Exploring Data
Let's explore the data a little.
```{r, cache=TRUE}
str(training)
head(training, n = 3)
```

# Training Set Partitioning
We divide the dataset in two set: 70% for myTraining and 30% for myTesting.
```{r, cache=TRUE}
inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
myTraining <- training[inTrain,] 
myTesting <- training[-inTrain,]
```

Let's see how much data we have in each set.
```{r, cache=TRUE}
dim(myTraining)
dim(myTesting)
```

# Data Cleaning
We perform some data transormation to ease the model training phase.

### Cleaning Near Zero Variance Variables
We remove from the dataset the columns that have a very low variance.
```{r, cache=TRUE}
nzv <- nearZeroVar(myTraining)
myTraining <- myTraining[, -nzv]
myTesting <- myTesting[, -nzv]
```

### Cleaning NA Variables
We remove from the dataset the variables that are mostly NAs.
```{r, cache=TRUE}
mostlyNA <- sapply(myTraining, function(x) mean(is.na(x))) > 0.95
myTraining <- myTraining[, mostlyNA==F]
myTesting <- myTesting[, mostlyNA==F]
```

### Remove the first five columns
The first five columns (X, user_name, raw_timestamp_part_1, raw_timestamp_part_2, cvtd_timestamp) are intuitively not useful for prediction, so we remove them.
```{r, cache=TRUE}
myTraining <- myTraining[, -(1:5)]
myTesting <- myTesting[, -(1:5)]
```

We reduced the number of prediction variables from `r dim(training)[[2]]` to `r dim(myTraining)[[2]]`.

# Using ML algorithms for prediction: Random Forest
We use [Random Forest](https://en.wikipedia.org/wiki/Random_forest) algorithm to train our model.
We use 3-fold cross-validation to select the optimal parameters.

```{r, cache=TRUE, message=FALSE, warning=FALSE}
fitControl <- trainControl(method="cv", number=3, verboseIter=F) # 3-fold CV 
fit <- train(classe ~ ., data=myTraining, method="rf", trControl=fitControl) # fit the model 
```

Let's print the final model and see the parameters
```{r, cache=TRUE}
fit$finalModel
```

# Model evaluation
Let's see how the trained model perform over our test set.
```{r, cache=TRUE}
preds <- predict(fit, newdata=myTesting) # make prediction in the validation set
confusionMatrix(myTesting$classe, preds) # show confusion matrix
```

The accuracy is very high (`r round(confusionMatrix(myTesting$classe, preds)[[3]][[1]]*100,1)`%) and the estimated "out of sample error" (error rate you get on new data set) is 0.3%.

Given the high accuracy we will use the "Random Forest model" to predict on the test set.

# Re-training the Model
Before that, we re-train the model using the full dataset to produce most accurate prediction.
We repeat all the steps above.

```{r, cache=TRUE}
# Cleaning Near Zero Variance Variables
nzv <- nearZeroVar(training)
training <- training[, -nzv]
testing <- testing[, -nzv]

# Cleaning NA Variables
mostlyNA <- sapply(training, function(x) mean(is.na(x))) > 0.95
training <- training[, mostlyNA==F]
testing <- testing[, mostlyNA==F]

# Remove the first five columns
training <- training[, -(1:5)]
testing <- testing[, -(1:5)]

# re-fit model
fitControl <- trainControl(method="cv", number=3, verboseIter=F)
fit <- train(classe ~ ., data=training, method="rf", trControl=fitControl)
```

# Prediction on Test Set 
Now, we use the model to predict the label for the test set.
```{r, cache=TRUE}
preds <- predict(fit, newdata=testing)
print(preds)
```

# Conclusions
We loaded the dataset and performed a little exploration of it.
We then divided the dataset in two sets: Training and Test set.
We then cleaned the dataset to reduce the number of predictor variables.
Subsequently we trained a model using the Random Forest Algorithm achieving a precision of 99.7% and an estimated out of sample error of 0.3%.
Finally we re-trained the model using the complete dataset and used it to predict the label of the test set.